---
title: "R Notebook"
output: html_notebook
---
# Load necessary library
```{r}
library(ggplot2)
library(timeDate)
library(dplyr)
library(caret)
library(e1071)
library(randomForest)
```

# Set up relevant directories
```{r}
wdir="E:\\R\\energyConsumptionForecast"
setwd(wdir)
```

# Load the data
```{r}
energyData <- read.csv("energy-consumption1.csv", skip = 2, header = FALSE, stringsAsFactors = FALSE)

head(energyData)
energyLabels <- "e0"
for (iii in 1:47){
  energyLabels = c(energyLabels, paste("e", as.character(iii), sep = ""))
}
colnames(energyData) <- c("Location", "Utility", "Unit", "Date", energyLabels, "Sum")
energyData <- energyData %>% filter(Location == "Kew site")
energyData$Date <- as.Date(energyData$Date, format = "%d/%m/%Y")
```

# Check the consistency

```{r}
energyData %>% filter(is.na(Date))
# not needed anymore
#energyData <- energyData %>% filter(!is.na(Date))
min(energyData$Date)
max(energyData$Date)
```

# Define the function that transposes the data into simple index-value form

```{r}
transpose <- function(x, locData, energyLabels){
  rowDF <- data.frame(as.double(x[energyLabels]))
  locData$tmptmp <- x
  colnames(rowDF) <- c("Energy")
  rowDF$Energy <- as.double(rowDF$Energy)
  locData$dataFrame <- rbind(locData$dataFrame, rowDF, make.row.names = TRUE)
}
```

# Get the one year (or any other) slice of the data

```{r, eval=FALSE, include=FALSE}
locData <- new.env()

locData$dataFrame <- data.frame(Energy = as.double())
str(locData$dataFrame)
head(locData$dataFrame)

# one year of data
energySubset <- energyData %>% filter(Date >= as.Date("2012-01-01"), Date < as.Date("2013-01-01"))
```

# Execute the transposition

```{r, include=FALSE}
#supress output
sink("NUL")
apply(energySubset, 1, transpose, locData, energyLabels)
sink()
```

# Add time

```{r}
timeAxis <- seq(from = as.POSIXlt(min(energySubset$Date), tz = "GMT"), by = 1800, length.out = nrow(locData$dataFrame))
locData$dataFrame$Time <- timeAxis
head(locData$dataFrame)
tail(locData$dataFrame)
```

# Aggregate to one-hour resolution

```{r}
locData$dataFrame <- locData$dataFrame %>% group_by(format(Time, "%Y-%m-%d %H")) %>% summarise(mean(Energy), min(Time))
colnames(locData$dataFrame) <- c("hour", "Energy", "Time")
locData$dataFrame$hour <- NULL
head(locData$dataFrame, n = 20)
```

# Any problems with the data?
```{r}
locData$dataFrame %>% filter(is.na(Energy))
```

# How the data antually looks like?
```{r}
# plot all data
ggplot(data = locData$dataFrame, aes(x = Time, y = Energy)) + geom_line()
```

# Determine the seasonality in the simplest possible way: EWMA fiter

```{r}
nL <- 24*28
locData$dataFrame$season <- EMA(locData$dataFrame$Energy, n = nL)
locData$dataFrame$season[1:nL] <- locData$dataFrame$season[nL+1]
```

# Plot the data again
```{r}
ggplot(data = locData$dataFrame) + geom_line(aes(x = Time, y = Energy), col = "black") + geom_line(aes(x = Time, y = season), col = "red")
```

# Determine the seasons - turn into categorical variable

```{r}
nBins <- 3
locData$dataFrame$seasonLevels <- cut(locData$dataFrame$season, breaks = nBins, labels = seq(1, nBins))
head(locData$dataFrame, n = 10)
```

```{r}
ggplot(data = locData$dataFrame) + geom_line(aes(x = Time, y = Energy, colour = "Raw data")) + geom_line(aes(x = Time, y = season, colour = "Trend")) + geom_line(aes(x = Time, y = as.numeric(seasonLevels) * max(locData$dataFrame$Energy) / max(as.numeric(seasonLevels)), colour = "Season")) +
  scale_y_continuous(sec.axis = sec_axis(~ . * max(as.numeric(locData$dataFrame$seasonLevels)) / max(locData$dataFrame$Energy), name = "Season level")) +
  labs(y = "Energy", x = "Date and time", colour = "Curves")
```

# Create predictors
```{r}
# create day of week feature
locData$dataFrame$dayOfWeek <- as.factor(dayOfWeek(timeDate(locData$dataFrame$Time)))
# create hour feature
locData$dataFrame$hour <- as.factor(format(locData$dataFrame$Time, "%H"))
head(locData$dataFrame)
```

# Convert to "othogonal" space

```{r}
dataset <- data.frame(predict(dummyVars(~ dayOfWeek + hour + seasonLevels + Energy + Time, data = locData$dataFrame), newdata = locData$dataFrame))
dataset$Time <- as.POSIXct(dataset$Time, origin = "1970-01-01", tz = "GMT")
head(dataset, n = 24)
```

# Create training index
```{r}
trainindex <- which(dataset$Time < as.POSIXct("2012-11-01 00:00", format = "%Y-%m-%d %H:%M", tz = "GMT") | dataset$Time >= as.POSIXct("2012-12-01 00:00", format = "%Y-%m-%d %H:%M", tz = "GMT"))
```

# Create feature-response vector
```{r}
#7 (dayOfWeek) + 24 (hour) + 1 (season) + 1 (response)
featuresResponseCols <- c(seq(1, 7), seq(8, 8+23), seq(32, 34), 35)
#featuresResponseCols <- c(seq(1, 7), seq(8, 8+23), 33)
NdataSetCols <- length(featuresResponseCols)

head(dataset[trainindex, featuresResponseCols])
```

# Get the training set
```{r}
training <- as.data.frame(dataset[trainindex, featuresResponseCols])
rownames(training) = NULL
head(training)
```

# Get the testing set
```{r}
testing = as.data.frame(dataset[-trainindex, featuresResponseCols])
rownames(testing) = NULL
head(testing)
```

# Train the model
```{r}
type <- "eps-regression" #regression
u <- -2 # -3,-2,-1,0,1,2,3
gam <- 10^{u} 
w= 4.5 #1.5,-1,0.5,2,3,4
cost <- 10^{w}

# support vector machine
svmFit <- svm(training[,1:(NdataSetCols-1)], 
              training[,NdataSetCols], 
              type = type,
              kernel = "radial",
              gamma = gam,
              cost = cost)
summary(svmFit)
```

# Predict - this is actually forecasting step
```{r}
predsvm <- predict(svmFit, testing[,1:(NdataSetCols-1)])
```

# Prepare visualisation data frame
```{r}
actualTS <- testing[,NdataSetCols]
predicTS <- predsvm

plotData <- data.frame(actualTS, predicTS)
colnames(plotData) <- c("real", "predicted")
plotData$time <- dataset[-trainindex, "Time"]
plotData$delta <- plotData$predicted - plotData$real
plotData$dow <- dayOfWeek(timeDate(plotData$time))
head(plotData)
```

# Plot the results
```{r Comparison between real and predicted values}
ggplot(data = plotData) + geom_line(aes(x = time, y = real, color = "Real values")) + 
  geom_line(aes(x = time, y = predicted, color = "Predicted values")) + 
  theme(legend.position = "right") +
  labs(y = "Energy", x = "Date and time (forecasted period)", colour = "Curves")
```

