---
title: "Forecasting of time series with the ML approaches"
output: html_notebook
---
# Change local settings
```{r}
Sys.setenv("LANGUAGE"="En")
Sys.setlocale("LC_ALL", "English")
```

# Load necessary library
```{r, echo = FALSE}
library(ggplot2)
library(timeDate)
library(dplyr)
library(caret)
library(e1071)
library(randomForest)
library(TTR)
library(xts)
library(forecast)
```
# Set up relevant directories
```{r}
wdir="E:\\R\\energyConsumptionForecast"
setwd(wdir)
```

# Load the data
```{r}
energyData <- read.csv("energy-consumption1.csv", skip = 2, header = FALSE, stringsAsFactors = FALSE)

head(energyData)
energyLabels <- "e0"
for (iii in 1:47){
  energyLabels = c(energyLabels, paste("e", as.character(iii), sep = ""))
}
colnames(energyData) <- c("Location", "Utility", "Unit", "Date", energyLabels, "Sum")
energyData <- energyData %>% filter(Location == "Kew site")
energyData$Date <- as.Date(energyData$Date, format = "%d/%m/%Y")
```

# Time period covered
```{r}
print(paste("Time interval covered by measures data:", min(energyData$Date), "-", max(energyData$Date), sep = " "))
```

# Define the function that transposes the data into simple index-value form
```{r}
transpose <- function(x, locData, energyLabels){
  rowDF <- data.frame(as.double(x[energyLabels]))
  locData$tmptmp <- x
  colnames(rowDF) <- c("Energy")
  rowDF$Energy <- as.double(rowDF$Energy)
  locData$dataFrame <- rbind(locData$dataFrame, rowDF, make.row.names = TRUE)
}
```

# Get the one year (or any other) slice of the data

```{r, result = "hide"}
locData <- new.env()

locData$dataFrame <- data.frame(Energy = as.double())
#str(locData$dataFrame)
#head(locData$dataFrame)

# one year of data
energySubset <- energyData %>% filter(Date >= as.Date("2012-01-01"), Date < as.Date("2013-01-01"))
```

# Execute the transposition
```{r, include=FALSE}
#supress output
sink("NUL")
apply(energySubset, 1, transpose, locData, energyLabels)
sink()
```

# Add time
```{r}
timeAxis <- seq(from = as.POSIXlt(min(energySubset$Date), tz = "GMT"), by = 1800, length.out = nrow(locData$dataFrame))
locData$dataFrame$Time <- timeAxis
head(locData$dataFrame)
```

# Aggregate to one-hour resolution
```{r}
locData$dataFrame <- locData$dataFrame %>% group_by(format(Time, "%Y-%m-%d %H")) %>% summarise(mean(Energy), min(Time))
colnames(locData$dataFrame) <- c("hour", "Energy", "Time")
locData$dataFrame$hour <- NULL
head(locData$dataFrame, n = 20)
```

# Any problems with the data?
```{r}
locData$dataFrame %>% filter(is.na(Energy))
locData$dataFrame <- locData$dataFrame %>% filter(!is.na(Energy) & !is.na(Time))
head(locData$dataFrame)
```

# How the data actually looks like?
```{r}
# plot all data
ggplot(data = locData$dataFrame, aes(x = Time, y = Energy)) + geom_line()
```

# Seasonal decomposition
```{r}
seasonalTS <- msts(locData$dataFrame$Energy, seasonal.periods = c(24*7))
decomposedTS <- decompose(seasonalTS)

decomposedDF <- data.frame(decomposedTS[1], decomposedTS[2], decomposedTS[3], decomposedTS[4], Time = locData$dataFrame$Time)
decomposedDF <- decomposedDF %>% filter(!is.na(trend))
#head(decomposedDF)
#head(locData$dataFrame$Energy)



ggplot(data = decomposedDF) + geom_line(aes(x = Time, y = x, colour = "Raw data")) +
                              geom_line(aes(x = Time, y = seasonal, colour = "Seasonality")) +
                              geom_line(aes(x = Time, y = trend, colour = "Trend"), size = 1.1) +
                              geom_line(aes(x = Time, y = random, colour = "Random")) +
                              labs(y = "Energy", x = "Date and time", colour = "Curves") +
                              scale_y_continuous() + ggtitle("Decomposition into trend, seasonality and random componentes") +
                              theme(legend.position="right",
                              legend.text = element_text(size=9),
                              axis.text = element_text(size=9),
                              axis.title = element_text(size=9),
                              plot.title=element_text(size = 9),
                              legend.title=element_text(size=9)) + scale_colour_manual(values=c("red","green","blue", "black"))
```
# Dicrete level creation
```{r}
# 1. cutting season into dicrete space
minE <- min(decomposedDF$trend)
maxE <- max(decomposedDF$trend)
deltaE <- maxE - minE

decomposedDF$seasonLevels <- cut(decomposedDF$trend, breaks = c(minE - 1, minE + 0.28*deltaE, minE + 0.75*deltaE, maxE), labels = seq(1, 3))

# 2. visualisation
ggplot(data = decomposedDF) + geom_line(aes(x = Time, y = x, colour = "Raw data")) +
                              geom_line(aes(x = Time, y = trend, colour = "Trend"), size = 1.1) + 
                              geom_line(aes(x = Time, y = as.numeric(decomposedDF$seasonLevels) * max(decomposedDF$x) / max(as.numeric(decomposedDF$seasonLevels)), colour = "Season level"), linetype = "solid", size = 1.1) +
                              scale_y_continuous(sec.axis = sec_axis(~ . * max(as.numeric(decomposedDF$seasonLevels)) / max(decomposedDF$x), name = "Season level")) +
                              labs(y = "Energy", x = "Date and time", colour = "Curves") +
                              theme(legend.position="right",
                              legend.text = element_text(size=9),
                              axis.text = element_text(size=9),
                              axis.title = element_text(size=9),
                              plot.title=element_text(size = 9),
                              legend.title=element_text(size=9)) + scale_colour_manual(values=c("green","red","black"))
```

```{r}
# 3. Expert-activity:  removal of redundant periods
maxSeasones <- max(as.numeric(decomposedDF$seasonLevels))

seasonLevelHigh <- (decomposedDF %>% filter(seasonLevels == 3, as.Date(Time) <= as.Date("2012-12-31")))
decomposedDF[decomposedDF$Time >= min(seasonLevelHigh$Time) & decomposedDF$Time <= max(seasonLevelHigh$Time), "seasonLevels"] <- 3
seasonLevelLow <- (decomposedDF %>% filter(seasonLevels == 1, Time < as.POSIXct("2012-06-01 00:00", format = "%Y-%m-%d %H:%M")))
decomposedDF[decomposedDF$Time >= min(seasonLevelLow$Time) & decomposedDF$Time <= max(seasonLevelLow$Time), "seasonLevels"] <- 2

ggplot(data = decomposedDF) + geom_line(aes(x = Time, y = x, colour = "Raw data")) +
                              geom_line(aes(x = Time, y = trend, colour = "Trend"), size = 1.1) + 
                              geom_line(aes(x = Time, y = as.numeric(decomposedDF$seasonLevels) * max(decomposedDF$x) / max(as.numeric(decomposedDF$seasonLevels)), colour = "Season level"), linetype = "solid", size = 1.1) +
                              scale_y_continuous(sec.axis = sec_axis(~ . * max(as.numeric(decomposedDF$seasonLevels)) / max(decomposedDF$x), name = "Season level")) +
                              labs(y = "Energy", x = "Date and time", colour = "Curves") +
                              theme(legend.position="right",
                              legend.text = element_text(size=9),
                              axis.text = element_text(size=9),
                              axis.title = element_text(size=9),
                              plot.title=element_text(size = 9),
                              legend.title=element_text(size=9)) + scale_colour_manual(values=c("green","red","black"))
```
# Create predictors
```{r}
# create day of week feature
decomposedDF$dayOfWeek <- as.factor(dayOfWeek(timeDate(decomposedDF$Time)))
# create hour feature
decomposedDF$hour <- as.factor(format(decomposedDF$Time, "%H"))
head(decomposedDF)
```

# Convert to "othogonal" space, "dummification"" of categorical variables
```{r}
DVobject <- dummyVars(~ dayOfWeek + hour + seasonLevels + x + Time, data = decomposedDF)
dataset <- as.data.frame(predict(DVobject, newdata = decomposedDF))
dataset$Time <- as.POSIXct(dataset$Time, origin = "1970-01-01", tz = "GMT")
head(dataset)
```

# Create training index
```{r}
trainindex <- which(dataset$Time < as.POSIXct("2012-02-01 00:00", format = "%Y-%m-%d %H:%M", tz = "GMT") | 
                    dataset$Time >= as.POSIXct("2012-03-01 00:00", format = "%Y-%m-%d %H:%M", tz = "GMT"))
```

# Create feature-response vector
```{r}
# 7 (dayOfWeek) + 24 (hour) + 3 (season) + 1 (response)
head(dataset, n = 1)
featuresResponseCols <- c(1:7, 8:(8+23), (8+23+1):(8+23+1+2), 35)
NdataSetCols <- length(featuresResponseCols)
head(dataset[trainindex, featuresResponseCols])
```

# Get the training set
```{r}
training <- as.data.frame(dataset[trainindex, featuresResponseCols])
rownames(training) = NULL
head(training)
```

# Get the testing set
```{r}
testing = as.data.frame(dataset[-trainindex, featuresResponseCols])
rownames(testing) = NULL
head(testing)
```

# Train the model: Support Vector Machines
[Support Vector Machines](http://www.kdnuggets.com/2016/07/support-vector-machines-simple-explanation.html)
```{r}
type <- "eps-regression" #regression
u <- -2 # -3,-2,-1,0,1,2,3
gam <- 10^{u} 
w <- 4.5 #1.5,-1,0.5,2,3,4
cost <- 10^{w}

# support vector machine
svmFit <- svm(training[,1:(NdataSetCols-1)], 
              training[,NdataSetCols], 
              type = type,
              kernel = "radial",
              gamma = gam,
              cost = cost)
summary(svmFit)
```

# Predict - this is actually forecasting step
```{r}
predsvm <- predict(svmFit, testing[, 1:(NdataSetCols-1)])
```

# Prepare visualisation data frame
```{r}
actualTS <- testing[,NdataSetCols]
predicTS <- predsvm

plotData <- data.frame(actualTS, predicTS)
colnames(plotData) <- c("real", "predictedSVM")
plotData$Time <- dataset[-trainindex, "Time"]
head(plotData)
```

# Plot the results
```{r Comparison between real and predicted values}
ggplot(data = plotData) + geom_line(aes(x = Time, y = real, color = "Real values")) + 
  geom_line(aes(x = Time, y = predictedSVM, color = "Predicted values: SVM")) + 
  theme(legend.position = "right") +
  labs(y = "Energy", x = "Date and time (forecasted period)", colour = "Curves") +
  scale_colour_manual(values=c("red","blue"))
```

# Train the model: Random Forest
[Random Forest](http://www.kdnuggets.com/2016/09/decision-trees-disastrous-overview.html)
```{r}
rfFit <- randomForest(training[,1:(NdataSetCols-1)], 
                      training[,NdataSetCols], ntree=100)
summary(rfFit)
```

```{r}
predrf <- predict(rfFit, testing[,1:(NdataSetCols-1)])
```
# Prepare visualisation data frame
```{r}
plotData$predictedRF <- predrf

ggplot(data = plotData) + geom_line(aes(x = Time, y = real, color = "Real values")) + 
  geom_line(aes(x = Time, y = predictedSVM, color = "Predicted values: SVM")) +
  geom_line(aes(x = Time, y = predictedRF, color = "Predicted values: RF")) + 
  theme(legend.position = "right") +
  labs(y = "Energy", x = "Date and time (forecasted period)", colour = "Curves") +
  scale_colour_manual(values=c("red","green","blue"))
```